---
title: "ECDF and Mahalanobis Distance for Empirical Niche Modeling"
author: "Luíz Fernando Esser"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ecdfniche_comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Introduction

This vignette shows how to use the **ECDFniche** package to reproduce the simulations from the original manuscript, comparing Mahalanobis distance–based suitability transformations using the chi-squared distribution and the empirical cumulative distribution function (ECDF).

We follow a *virtual ecologist* approach (Zurell et al. 2010) to evaluate how different transformations of the Mahalanobis distance recover a simulated niche across varying dimensionalities (number of predictor variables) and sample sizes (number of records), and then extend the analysis to a bivariate non-normal environmental space.

```{r}
library(ECDFniche)
```

## Multivariate normal niche: `ecdf_compare_niche()`

In the first set of simulations, we assume that the environmental predictors describing a species’ niche follow a multivariate normal distribution.

Dimensions \(p\) range from 1 to 5, and sample sizes \(n\) range from 20 to 500 in steps of 20, mimicking increasing numbers of occurrence records.

For each combination of \(p\) and \(n\), `ecdf_compare_niche()`:

1. Draws a fixed covariance matrix \(\Sigma\) by sampling from a Wishart distribution with \(p + 2\) degrees of freedom and scale matrix set to zero with a diagonal of 1, ensuring a positive-definite covariance for that \((p, n)\) combination.
2. Generates 30 independent replicates of \(n\) records from a \(p\)-variate normal distribution with mean vector \(\mu_p = \mathbf{0}\) (interpreted as the environmental optimum) and covariance matrix \(\Sigma\).
3. For each replicate, estimates the sample mean \(\mu_r\) and sample covariance \(\Sigma_r\) and computes the squared Mahalanobis distance
\[
D^2 = (x - \mu_r)^\top \Sigma_r^{-1} (x - \mu_r).
\]
4. Computes two suitability metrics for every record:
   - Theoretical chi-squared suitability:
\[
S_{\chi^2} = 1 - F_{\chi^2_p}(D^2),
\]
   where \(F_{\chi^2_p}\) is the cumulative distribution function of the chi-squared distribution with \(p\) degrees of freedom.
   
   - Empirical ECDF-based suitability:
\[
S_{\text{ECDF}} = 1 - F_{\text{ECDF}}(D^2),
\]
   where \(F_{\text{ECDF}}\) is the empirical cumulative distribution function of the distances.
   
5. Calculates Pearson’s correlation coefficient between \(S_{\chi^2}\) and \(S_{\text{ECDF}}\) for each replicate.

This setup mimics a realistic ENM scenario where multivariate means and true covariances are unknown and must be estimated from occurrence records drawn from a correlated environmental space representing the species’ theoretical niche (sensu Hutchinson 1978).

```{r}
set.seed(1991)
normal_res <- ecdf_compare_niche(
  p_vals = 1:5,
  n_vals = seq(20L, 500L, 20L),
  n_reps = 30L
)
```

### Figure 1: Correlation vs sample size

`cor_plot` summarizes, for each \(p\) and \(n\), the mean and standard deviation of the correlation between chi-squared and ECDF suitabilities across the 30 replicates, with individual replicate values shown as grey points. The plot shows the correlation between suitability metrics estimated using the chi-squared and the Empirical Cumulative Distribution Function (ECDF) across sample sizes and numbers of environmental predictors. Each panel shows Pearson correlation coefficients as a function of the number of occurrence records for a given dimensionality p (1 to 5 variables). Correlations are generally very high (rarely < 0.95), increasing with sample size and slightly increasing with dimensionality.

```{r, fig.width=11, fig.height=4}
normal_res$cor_plot
```

### Figure 2: Suitability vs Mahalanobis distance

`suit_plot` pools observations across sample sizes for each \(p\), recomputes an ECDF-based suitability on the combined distances, and compares these to chi-squared curves. The plot shows the relationship between squared Mahalanobis distance (x-axis) and environmental suitability metrics (y-axis) estimated using the chi-squared and the Empirical Cumulative Distribution Function (ECDF) across different numbers of environmental predictors. Grey curves represent the habitat suitability based on the chi-squared distribution while red circles represent habitat suitability via ECDF, highlighting that ECDF closely tracks the theoretical chi‑squared mapping over the distance range.

```{r, fig.width=11, fig.height=4}
normal_res$suit_plot
```

## Non-normal bivariate niche: `ecdf_nonnormal_niche()`

In many real ENM applications, environmental covariates are not jointly normal and species can show complex responses to environmental gradients (Anderson et al. 2022).

To explore this, `ecdf_nonnormal_niche()` simulates a bivariate environmental space for temperature and precipitation using a Gaussian copula:

- Temperature follows a normal distribution with mean 20 °C and standard deviation 5 °C.
- Precipitation follows a Weibull distribution with shape 2 and scale 10 mm, representing skewed rainfall data.
- The dependence between the two variables is controlled by a correlation parameter \(\rho \in \{-0.7, -0.3, 0, 0.3, 0.7\}\), reflecting negative to positive associations observed in climatological studies (e.g. Anderson et al. 2019).

For each \(\rho\), the function:

1. Uses a Gaussian copula with correlation \(\rho\) to generate a large reference sample of size \(N_{\text{ref}}\) (default \(10^5\)) and derive “true” population mean vector and covariance matrix for the bivariate distribution.
2. For each combination of \(\rho\) and sample size \(n \in \{20, 50, 100, 200, 500\}\), draws \(n_{\text{reps}}\) replicate samples directly from the copula-based bivariate non-normal distribution.
3. Computes squared Mahalanobis distances \(D^2\) using the *true* mean and covariance from the reference sample, isolating the effect of non-normality from parameter estimation error.
4. Calculates:
   - Chi-squared suitability \(S_{\chi^2} = 1 - F_{\chi^2_2}(D^2)\), which is theoretically “wrong” under non-normality but serves as the standard parametric benchmark.
   - ECDF-based suitability \(S_{\text{ECDF}} = 1 - F_{\text{ECDF}}(D^2)\).
5. Returns per-replicate correlations and observation-level data, plus a summary plot comparing the two suitabilities.

```{r}
set.seed(1991)
nonnormal_res <- ecdf_nonnormal_niche(
  rho_vals = c(-0.7, -0.3, 0, 0.3, 0.7),
  n_vals   = c(20L, 50L, 100L, 200L, 500L),
  n_reps   = 10L,
  N_ref    = 1e5,
  temp_function = "qnorm",
  temp_parameters = list(mean = 20, sd = 5),
  prec_function = "qweibull",
  prec_parameters = list(shape = 2, scale = 10)
)
```

### Figure 3: Suitability vs distance under non-normality

`suit_plot` shows ECDF-based suitability (colored by \(\rho\)) and chi-squared suitability (red points) as functions of \(D^2\), faceted by sample size. Plots highlight the sensitivity of chi-squared and ECDF suitability metrics to sample size and variable correlation in non-normal bivariate data. Internal histograms represent the distribution of ECDF-based values around the chi-squared-based suitability. The ECDF estimator shows higher stochasticity in small samples but converges to the chi-squared expectation in larger samples.


```{r, fig.width=11, fig.height=4}
nonnormal_res$suit_plot
```

## Customizing simulations

You can customize key aspects of both simulations to reproduce or extend the analyses:

- **Multivariate normal niche** (`ecdf_compare_niche()`):
  - `p_vals`: set of dimensions (e.g. `1:10`) to evaluate high-dimensional behavior.
  - `n_vals`: sample size grid, e.g. smaller `n` to explore the effect of limited records.
  - `n_reps`: number of replicates per combination for more stable summaries.

```{r, fig.width=11, fig.height=3}
res_custom_normal <- ecdf_compare_niche(
  p_vals = 2:4,
  n_vals = seq(20L, 200L, 20L),
  n_reps = 50L,
  seed   = 42
)
res_custom_normal$cor_plot
```

- **Non-normal bivariate niche** (`ecdf_nonnormal_niche()`):
  - `rho_vals`: alternative dependence structures.
  - `n_vals`, `n_reps`: sample size and replication design.
  - `N_ref`: reference population size; larger values approximate the true parameters more closely.

```{r, fig.width=11, fig.height=3}
res_custom_nonnorm <- ecdf_nonnormal_niche(
  rho_vals = c(-0.5, 0, 0.5),
  n_vals   = c(30L, 100L, 300L),
  n_reps   = 20L,
  N_ref    = 5e4,
  temp_function = "qnorm",
  temp_parameters = list(mean = 20, sd = 5),
  prec_function = "qweibull",
  prec_parameters = list(shape = 2, scale = 10),
  seed     = 123
)
res_custom_nonnorm$suit_plot
```

Together, these simulations illustrate when the traditional chi-squared transformation is reliable and when the ECDF-based approach provides a more robust mapping from niche-based distances to habitat suitability, particularly when environmental covariates deviate from multivariate normality.
